Functional Enrichment Tests: Intro and methods comparison

### 2023-07-06 DRAFT_Funct_Enrich.Rmd
SPID to go

@@

In a majority of cases you will want to annotate a gene set to get gene ontology information. If you are creating your own genome or transcriptome it should be rather straightforward to know what file to annotate. If using a widely studied system where there are publically available resources, it is advisable to use those as this is the best way to facilitate integration of data sets. In this case study we will be considering the Eastern oyster, (_Crassostrea virginica_) for which there is data at [NCBI](https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_002022765.2/) and [Ensembl Metazoa](http://metazoa.ensembl.org/Crassostrea_virginica_gca002022765v4/Info/Index). At NCBI there is both a GenBank and RefSeq assembly available.

@@

@@

```
>lcl|NC_035780.1_prot_XP_022327646.1_1 [gene=LOC111126949] [db_xref=GeneID:111126949] [protein=UNC5C-like protein] [protein_id=XP_022327646.1] [location=join(30535..31557,31736..31887,31977..32565,32959..33204)] [gbkey=CDS]
MTEVCYIWASSSTTVVICGIFFIVWRCFISIKKRASPLHGSSQQVCQTCQIEGHDFGEFQLSCRRQNTNVGYDLQGRRSD

```
```{r engine='bash', eval=FALSE}
/home/shared/ncbi-blast-2.11.0+/bin/blastp \
-query DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.faa \
-db DRAFT_Funct_Enrich/annot/uniprot_sprot_r2023_02 \
-out DRAFT_Funct_Enrich/annot/Cvir_transcds-uniprot_blastp.tab \
-evalue 1E-20 \
-num_threads 40 \
-max_target_seqs 1 \
-outfmt 6
```

@@

```{r engine='bash', eval=TRUE}
head -2 DRAFT_Funct_Enrich/annot/Cvir_transcds-uniprot_blastp.tab \
```

@@

@@

!tr '|' '\t' < $blastout \
> _blast-sep.tab

!awk -v OFS='\t' '{print $4, $1, $14}' < _blast-sep.tab | sort \
> _blast-sort.tab

>>>>

```{r}
library(tidyverse)
```

@@

In a majority of cases you will want to annotate a gene set to get gene ontology information. If you are creating your own genome or transcriptome it should be rather straightforward to know what file to annotate. If using a widely studied system where there are publically available resources, it is advisable to use those as this is the best way to facilitate integration of data sets. In this case study we will be considering the Eastern oyster, (*Crassostrea virginica*) for which there is data at [NCBI](https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_002022765.2/) and [Ensembl Metazoa](http://metazoa.ensembl.org/Crassostrea_virginica_gca002022765v4/Info/Index). At NCBI there is both a GenBank and RefSeq assembly available.

@@

echo "number of sequences"
grep -c  ">" DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.faa

@@

```         
>lcl|NC_035780.1_prot_XP_022327646.1_1 [gene=LOC111126949] [db_xref=GeneID:111126949] [protein=UNC5C-like protein] [protein_id=XP_022327646.1] [location=join(30535..31557,31736..31887,31977..32565,32959..33204)] [gbkey=CDS]
MTEVCYIWASSSTTVVICGIFFIVWRCFISIKKRASPLHGSSQQVCQTCQIEGHDFGEFQLSCRRQNTNVGYDLQGRRSD
```

```{r engine='bash', eval=FALSE}
/home/shared/ncbi-blast-2.11.0+/bin/blastp \
-query DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.faa \
-db DRAFT_Funct_Enrich/annot/uniprot_sprot_r2023_02 \
-out DRAFT_Funct_Enrich/annot/Cvir_transcds-uniprot_blastp.tab \
-evalue 1E-20 \
-num_threads 40 \
-max_target_seqs 1 \
-outfmt 6
```

@@

```{r engine='bash', eval=TRUE}
head -2 DRAFT_Funct_Enrich/annot/Cvir_transcds-uniprot_blastp.tab
```

```{r}
blast <- read.csv("DRAFT_Funct_Enrich/annot/Cvir_transcds-uniprot_blastp.tab", sep = '\t', header = FALSE)
```

curl -O https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_genomic.fna.gz

gunzip -k GCF_002022765.2_C_virginica-3.0_genomic.fna.gz
```

@@

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("rtracklayer")

# Load the library
library(rtracklayer)

# Define the file path to your GFF file
file_path <- "DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_genomic.gff"

# Import the GFF file
gff_data <- import(file_path)

# Convert GRanges object to a DataFrame for easier manipulation
gff_data_frame <- as.data.frame(gff_data)
```

```{r}
cds <- gff_data_frame %>%
  filter(gbkey == "CDS", source == "Gnomon")
```

Convert fasta to tab

```{bash}
perl -e '$count=0; $len=0; while(<>) {s/\r?\n//; s/\t/ /g; if (s/^>//) { if ($. != 1) {print "\n"} s/ |$/\t/; $count++; $_ .= "\t";} else {s/ //g; $len += length($_)} print $_;} print "\n"; warn "\nConverted $count FASTA records in $. lines to tabular format\nTotal sequence length: $len\n\n";' \
DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.faa > DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.tab
```

```{bash}
head -1 DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.tab
```

```{r}
cdsftab <- read.csv("DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.tab", sep = '\t', header = FALSE, row.names=NULL)
```

join blast with cdsftab

```{r}
g.spid <- left_join(blast, cdsftab, by = "V1") %>%
  mutate(gene = str_extract(V2.y, "(?<=\\[gene=)\\w+")) %>%
  select(gene, V11, V2.x) %>%
  mutate(SPID = str_extract(V2.x, "(?<=\\|)[^\\|]*(?=\\|)")) %>%
  distinct(gene, SPID, .keep_all = TRUE)
```

```{r}
left_join(blast, cdsftab, by = "V1") %>%
  mutate(gene = str_extract(V2.y, "(?<=\\[gene=)\\w+")) %>%
  select(gene, V11, V2.x) %>%
  mutate(SPID = str_extract(V2.x, "(?<=\\|)[^\\|]*(?=\\|)")) %>%
  distinct(gene, SPID, .keep_all = TRUE) %>%
  select(SPID) %>%
  write.table(file = "DRAFT_Funct_Enrich/annot/SPID.txt", sep = "\t", row.names = FALSE, quote = FALSE
 ) 
```

```{python}
python DRAFT_Funct_Enrich/annot/uniprot-retrieval.py DRAFT_Funct_Enrich/annot/SPID.txt
```

@@

!tr '\|' '\t' \< \$blastout\
\> \_blast-sep.tab

!awk -v OFS='\t''{print \$4, \$1, \$14}' \< \_blast-sep.tab \| sort\
\> \_blast-sort.tab
